Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-01-20 Manuscript/manuscript.Rmd
Methods reorganized and rewritten

3. **Distance**: For each genomic bin, the log [??? didn't you test raw distance first?] [??? log10?] of the distance (in bases) from the center of the bin to the center of the nearest genomic feature region was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1)

To assess the predictive performance for each predictor type, we performed individual random forest classification algorithms utilizing each set predictor types as the feature space, in addition to a model with all combined types. Precision-Recall (PR) Curves, and subsequent area under the PR curves (AUPRC), were used to evaluate which predictor type contributed to the best model performance. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models within each set of predictor types. Here, random forest models were used to assign weights to a set features of a given size. The optimal number of features was determined by the AUPRC of the corresponding model.

## ENSEMBLE FRAMEWORK

An ensemble of models was established for the purposes of systematically comparing the performances of different parameters in the framework. Each system in the framework is assumed to be composed of a model constructed from a particular data resolution, with a particular predictor type for the feature space (Figure 2).

@@

We then evaluated three re-sampling techniques (in addition to no re-sampling) on the reduced training set, in order to evaluate the problem of imbalanced classes. They are detailed below:

@@

3. **Random Over-Sampling (ROS)**: All of the majority classes from the training set were used. Sampling with replacement was used to obtain the same number of the minority classes. Similar to the under-sampling technique, 50 iterations were performed and results were aggregated by taking the average across iterations.

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. We chose the parameters that would create perfectly balanced classes. The `DMwR` package in R (version 3.4.2) was used to perform SMOTE.

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernal density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` package in R (version 3.4.2) was used to perform ROSE.

@@

A classification algorithm in the form of a random forest was performed and validated using the same test set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. The evaluation metrics listed above are all considered single threshold measures. That is, they are defined for individual thresholds that separated class prediction. Thererfore they are unable to give an overview of the potential range of performance as thresholds vary. A full list of the single threshold measures used (including additional metrics not reported), along with their respective formulas, are provided in Supplementary Table 6. In order to evaluate the trade-off between precision and recall, we also considered a threshold free measure in form of the area under the precision recall curve (AUPRC).

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest model. For each tree, the prediction accuracy on the out-of-bag portion of the data is recorded. Then the same was done after permuting each predictor variable. The difference between the two accuracies were then averaged over all trees, and normalized by the standard error [@kuhn2012caret].

@@

We evaluated and compared Hi-C datasets at four different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}, used three predictor types P = {OC, OP, Distance}, two variable reduction techniques V = {LASSO, Elastic-Net}, and five class balancing techniques C = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {r, p, v, c}, where r $\in$ R, p $\in$ P, v $\in$ V, and c $\in$ C. Letting |X| denote the cardinality of each set, we evaluated |R| x |P| x |V| x |C| different ensemble systems, totaling 120 (=4 $\times$ 3 $\times$ 2 $\times$ 5) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables.

>>>>

3. **Distance**: For each genomic bin, the log [??? didn't you test raw distance first?] [??? log10?] of the distance (in bases) from the center of the bin to the center of the nearest genomic feature region was calculated. The log transformation was used as a normalization technique in order to account for the skewness of the data (Supplementary Figure 1).

@@

In order to evaluate the problem of imbalanced classes, we then evaluated three re-sampling techniques.

@@

3. **Random Over-Sampling (ROS)**: All of the majority classes from the training set were used. Sampling with replacement was used to obtain the same number of the minority classes. Similar to the under-sampling technique, 50 iterations were performed and the results of a random forest classification were aggregated by taking the average across iterations.

4. **SMOTE** (Synthetic Minority Over-Sampling Technique): This method incorporates both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class [@chawla2002smote]. To account for the bias introduced by the under-sampling, we performed 50-iterations of SMOTE. We chose the parameters that would create perfectly balanced classes. The `DMwR` v.3.4.2 R package was used to perform SMOTE.

5. **ROSE** (Random Over-Sampling Examples): Similar to SMOTE, ROSE combines techniques of over-sampling and under-sampling. However, contrary to SMOTE, all of the data generated by ROSE is synthetic. The data is created by sampling from one of the two classes, then generating a new example in its neighborhood using a smoothed bootstrap approach. The shape of the neighborhood is determined by the kernel density estimate of the two classes [@lunardon2013r]. Another important distinction between ROSE and SMOTE is that, for ROSE, the creation of perfectly balanced classes is unlikely due to the randomness in sampling between the two classes. The `ROSE` v.3.4.2 [??? Is it correct? Same version as DMwR?] R packagewas used to perform ROSE.

@@

A classification algorithm in the form of a random forest was performed and validated using the same test [??? Training?] set for all models in the ensemble. To reduce bias due to random dataset generation, 10-fold cross validation was used. The default number of features to consider at each node of the random forest algorithm was set as the square root of the number of features in the model. Likewise, the number of ensemble trees to aggregate was set at 500. Once the model was implemented, it was assessed using the test set, and performance metrics were then recorded.

@@

Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. The evaluation metrics listed above are all considered to be single threshold measures. That is, they are defined for individual thresholds that separated class prediction. Thererfore, they are unable to give an overview of the potential range of performance as thresholds vary. In order to evaluate the trade-off between precision and recall, we also considered a threshold-free measure in form of the area under the precision recall curve (AUPRC).

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest model. For each tree, the prediction accuracy on the out-of-bag portion of the data is recorded. Then the same was done after permuting each predictor variable. The difference between the two accuracies were then averaged over all trees, and normalized by the standard error [@kuhn2012caret]. Recursive feature elimination (RFE) was then used to determine the optimal number of features for models within each set of predictor types. Here, random forest models were used to assign weights to a set features of a given size. The optimal number of features was determined by the AUPRC of the corresponding model. [??? Rewrite this paragraph as I put some disjoint sentences]

@@

We created an ensemble framework to systematically evaluate the performance of of TAD boundary prediction under different conditions and resampling approaches (Figure 3). We evaluated and compared Hi-C datasets at four different resolutions R = {10 kb, 25 kb, 50 kb, 100 kb}, used three predictor types P = {OC, OP, Distance}, two variable reduction techniques V = {LASSO, Elastic-Net} [??? Is that so?], and five class balancing techniques C = {None, RUS, ROS, SMOTE, ROSE}. A system produced from the ensemble framework was defined as: E = {r, p, v, c}, where r $\in$ R, p $\in$ P, v $\in$ V, and c $\in$ C. Letting |X| denote the cardinality of each set, we evaluated |R| x |P| x |V| x |C| different ensemble systems, totaling 120 (=4 $\times$ 3 $\times$ 2 $\times$ 5) separate models. For simplicity, we present the most significant results, with the complete results available as Supplementary Tables.
